---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading and preprocessing the data

```{r loading and preprocessing}
setwd("~/Documents/DataScienceCoursera/Course 5/WD/") # set working directory
data <- read.csv("activity.csv") # read the data

str(data) # check data stucture and if there are NAs
data$date <- as.Date(data$date) # transform the date variable to class Date
```

### What is mean total number of steps taken per day?
```{r histogram}
stepsTotal <- as.data.frame(with(data, tapply(steps, date, sum, na.rm = TRUE))) # calculate the total number of steps for each day
names(stepsTotal) <- "Total_steps" # change the colname to a more legible name

## 1. Make a histogram of the total number of steps taken each day
hist(stepsTotal[,1], breaks = nrow(stepsTotal), col = "blue", main = "Total steps", xlab = "Number of steps per day") 

## create the directory figures if it does not exist
if (!file.exists("figures")){
        dir.create("figures")
}

## save the plot as a separate file
png(file="./figures/Histogram_total_steps.png")
hist(stepsTotal[,1], breaks = nrow(stepsTotal), col = "blue", main = "Total steps", xlab = "Number of steps per day") # generate histogram 
dev.off()


## 2. Calculate and report the **mean** and **median** total number of steps taken per day
## the mean and median is calculated for each day and values are plotted with bargraphs
stepsMean <- as.data.frame(with(data, tapply(steps, date, mean, na.rm = TRUE)))
data2 <- data[data$steps>0,]
stepsMedian <- as.data.frame(with(data2, tapply(steps, date, median, na.rm = TRUE)))

par(mfrow = c(1,2), mar = c(2, 4, 2, 1))
barplot(stepsMean[,1], main = "Total steps per day", ylab = "Mean")
barplot(stepsMedian[,1], main = "Total steps per day", ylab = "Median")

## save the plots as a separate file
png(file="./figures/mean_median_total_steps.png")
par(mfrow = c(1,2), mar = c(2, 4, 2, 1))
barplot(stepsMean[,1], main = "Total steps per day", ylab = "Mean")
barplot(stepsMedian[,1], main = "Total steps per day", ylab = "Median")
dev.off()
```

### What is the average daily activity pattern?

```{r}
## 1. Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

meanAlldays <- as.data.frame(with(data, tapply(steps, interval, mean, na.rm = TRUE)))
names(meanAlldays) <- "averageByInterval"
plot(unique(data$interval), meanAlldays[,1], type = "l", ylab = "Average number of steps", xlab = "time series (5-minute interval)")

## 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
maxInterval <- rownames(meanAlldays[which.max(meanAlldays[,1]),])

## save the plot as a separate file
png(file="./figures/average_daily_activity.png")
plot(unique(data$interval), meanAlldays[,1], type = "l", ylab = "Average number of steps", xlab = "time series (5-minute interval)")
dev.off()
```

The 5-minute interval that contains the maximum number of steps is 835.

### Imputing missing values

I am using a very simple stategy by assigning the mean values by interval calculated above to the respective missing values.

```{r}
## 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with `NA`s)
numberNA <- sum(is.na(data$steps)) 

## 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. 

data$ID <- rownames(data) # change to rownames IDs
meanAlldays$interval <- rownames(meanAlldays) # change to rownames IDs
missing <- data[is.na(data$steps),] # find the rows with missing values
fillMissing <- merge(meanAlldays, missing, by ="interval") # add the mean values per interval to each missing rows by merging by interval
tmp <- merge(data, fillMissing, by ="ID", all.x = TRUE) # add the new values to the origingal dataset
tmp$ID <- as.integer(tmp$ID) # change class

## replace the missing values only with the respective imputed data based on mean values by interval
for (i in 1:nrow(tmp)){
        if (is.na(tmp[i,2])){
                tmp[i,2]<- tmp[i,6]
        }
}

## 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
tmp <- tmp[order(tmp$ID),] # order the data 
rownames(tmp) <- tmp$ID # change the rownames
newData <- tmp[, c(1:4)] # select columns of interest
rm(tmp) # remove temporary variable
newData <- newData[, -c(1)] # remove unnecessary column
names(newData) <- c("steps", "date", "interval") # rename colnames of the new dataset


sum(is.na(newData$steps))  # check that there are no missing values

## 4. Make a histogram of the total number of steps taken each day and Calculate and report the **mean** and **median** total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

stepsTotalnew <- as.data.frame(with(newData, tapply(steps, date, sum)))
names(stepsTotalnew) <- "Total_steps imputed data"
hist(stepsTotalnew[,1], breaks = nrow(stepsTotalnew), col = "blue", main = "Total steps", xlab = "Number of steps per day")

## save the plot as a separate file
png(file="./total_steps_noNAs.png")
hist(stepsTotalnew[,1], breaks = nrow(stepsTotalnew), col = "blue", main = "Total steps", xlab = "Number of steps per day")
dev.off()

stepsMeannew <- as.data.frame(with(newData, tapply(steps, date, mean)))
data2new <- newData[newData$steps>0,]
stepsMediannew <- as.data.frame(with(data2new, tapply(steps, date, median)))

par(mfrow = c(2,2), mar = c(2, 4, 2, 1))
barplot(stepsMean[,1], main = "Total steps with NAs", ylab = "Mean per day")
barplot(stepsMedian[,1], main = "Total steps with NAs", ylab = "Median per day")
barplot(stepsMeannew[,1], main = "Total steps no NAs", ylab = "New Mean per day")
barplot(stepsMediannew[,1], main = "Total steps no NAs", ylab = "New Median per day")

## save the plot as a separate file
png(file="./figures/compareData_with_without_NAs.png")
par(mfrow = c(2,2), mar = c(2, 4, 2, 1))
barplot(stepsMean[,1], main = "Total steps with NAs", ylab = "Mean per day")
barplot(stepsMedian[,1], main = "Total steps with NAs", ylab = "Median per day")
barplot(stepsMeannew[,1], main = "Total steps no NAs", ylab = "New Mean per day")
barplot(stepsMediannew[,1], main = "Total steps no NAs", ylab = "New Median per day")
dev.off()
```

The number of missing values is `r numberNA`.

The imputation seems to have worked well and I dont see a signigficant change in the pattern of mean and median total steps per day between datasets with and without NAs.

### Are there differences in activity patterns between weekdays and weekends?

```{r, cache=TRUE}
## 1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
weekdays <- weekdays(newData$date)

## create a tag for each observational day
dayTag <-rep(NA, length(weekdays))
for(i in 1:length(weekdays)){
        if (weekdays[i] %in% unique(weekdays)[1:5]){
             dayTag[i] <- "weekday"   
        } else {
                dayTag[i] <- "weekend"
        }
}

newData$dayTag <- dayTag # assign the tag to a new column of the dataset

## subset the dataset based on the daytag
weekdayData <- subset(newData, dayTag =="weekday")
weekendData <- subset(newData, dayTag =="weekend")

## calculate the mean for both subsetted data
meanWeekday <- as.data.frame(with(weekdayData, tapply(steps, interval, mean)))
names(meanWeekday) <- "averageByInterval"
meanWeekday$Interval <- unique(newData$interval)
meanWeekday$dayTag <- rep("weekday")

meanWeekend <- as.data.frame(with(weekendData, tapply(steps, interval, mean)))
names(meanWeekend) <- "averageByInterval"
meanWeekend$Interval <- unique(newData$interval)
meanWeekend$dayTag <- rep("weekend")

## combine the two calculated sets of means by interval
meanByInterval <- rbind(meanWeekday, meanWeekend)

## 1. Make a panel plot containing a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
library(lattice)
xyplot(averageByInterval ~ Interval | dayTag, meanByInterval, type ="l", ylab = "Number of steps", layout = c(1,2))

## save the plot as a separate file
png(file="./figures/weekend_weekdays_activity.png")
xyplot(averageByInterval ~ Interval | dayTag, meanByInterval, type ="l", ylab = "Number of steps", layout = c(1,2))
dev.off()
```

